%% Section 03 — Background and Literature Review
\chapter{Background and Literature Review}
\label{ch:litreview}

\section{Introduction}
\label{sec:lr-intro}
Consensus is a collective decision-making process where every party can object to any
proposal and effectively veto it. It becomes a consensus regime when consensus becomes
the only means to reach decisions. Consensus is not the same as unanimity: a unanimity
rule puts a proposal to a vote and records who votes in favour, against or abstain, whereas under
consensus a decision proceeds whenever no party registers a formal objection \parencite{Rietig2023}. 
Since in such regimes a single party can halt any
proposal even when no other party objects, an outside observer might, at first sight,
expect such regimes to decide very little. This is even more pronounced when there is
a large and heterogeneous group: the larger the group, the smaller the set of proposals
that change the status quo \parencite{Tsebelis2002}. In practice, consensus regimes can reach agreement even on contested,
transnational issues, often by relying on informal practices: trust, dialogue, and leadership. These practices fill the gaps left by
their formal rules \parencite{Sykorabodie2019}. But success is never guaranteed: in bodies such as the United Nations Security Council
a single veto routinely blocks collective action, and even historically productive regimes can stall \parencite{Gardiner2025}. 
Quantitatively, the picture is no different: what separates a consensus regime that keeps reaching decisions from one that can
no longer reach them is not captured by existing models of consensus dynamics \parencite{Perez2018,Tran2024}. A model that better reproduces
how consensus is reached or broken is therefore needed, both to explain past behaviour and to anticipate when a
regime approaches the limits of its capacity to act.

Models of multi-actor decision-making already exist, but they were not built for this task. Classical
opinion-dynamics and game-theoretic models represent each actor with a single, low-dimensional position and are
rarely use the actual text those actors produce, such as their statements and negotiating documents. Capturing a consensus regime faithfully requires more: a
higher-dimensional description in which actors hold distinct positions across many issues, preferences read
directly from the text the actors produce and organised by topic, and measures validated and calibrated against
observed outputs rather than assumed. Recent advances in natural language processing and machine learning make
each of these upgrades feasible.

The Antarctic Treaty System (ATS) is one of those
consensus-driven regimes. At its centre is the Antarctic Treaty (AT), whose
decision-making parties are called consultative parties \parencite{ATS2023RoP}. These parties meet yearly at the Antarctic
Treaty Consultative Meeting (ATCM), where outputs are adopted only by consensus.
As international relations strain and Antarctic environmental pressures intensify,
the risk of losing the capacity to act increases. In 2022, for the first time, the
ATCM parties failed to reach consensus on the final report; even as activity has
increased, the regime has become less responsive to challenges \parencite{Gardiner2025}.


\section{Modelling consensus: responsive and anticipatory approaches}
\label{sec:consensus}
There are two traditional groups of models for how groups reach consensus. The first group is responsive and interactive: each actor repeatedly reacts to
the current positions of the others. These are the synchronization and opinion-dynamics models; examples are the Kuramoto and
opinion-changing-rate models used by \textcite{Botelho_CoalitionDynamics}, which represent agreement as a coupled-oscillator convergence. These methods treat consensus as \textit{convergence}. \textcite{DeGroot1974} iteratively average agents' opinions using \(x(t+1)=Wx(t)\) until they converge. A bounded-confidence version was later developed by \textcite{Hegselmann2002}.
A further model in this group is the stubborn-agents model \parencite{FriedkinJohnsen1990}, in which each agent only partially updates toward its neighbours and stays anchored to its initial opinion.
Opinion-dynamics models can nevertheless test for agreement: declare consensus when all final
positions lie within a tolerance radius, as \textcite{Bernardo2021} do for the negotiation of the
2015 Paris Agreement. That test, however, captures preference convergence, not the consensus rule
the ATS uses: a procedural right under which one party can block a decision no matter how close
the remaining positions are. These opinion-dynamics models also contain no strategic preference,
no veto, and no learning; the bargaining behaviour of consensus regimes is therefore outside
their scope.

The second group is anticipatory: each actor reasons about how the others will respond and
chooses its actions to maximise an objective function. Responsive agents, by contrast, follow an
update rule with no objective at all. Anticipatory models use social choice and game
theory to treat consensus as a unanimity rule. \textcite{Tsebelis2002} develop a veto-players framework in which every party has
veto power. Let $q$ denote the status quo, $p$ a proposal, and $u_i$ the utility function of
party $i$, which scores how much that party values each policy. The status quo is replaced only
when $u_i(p) \geq u_i(q)$ for every party $i$: no party may be left worse off than it is now.
Consequently, as the number of parties increases and their positions spread, the set of adoptable policies shrinks and the system
becomes harder to change. Analysing the ATS, \textcite{Gardiner2025} point to the same ingredients: a growing and
diversifying membership, increasingly divergent views, and a consensus rule, and view the
regime's shift away from binding agreements as symptomatic of declining decision-making
performance.

Each group has strengths and limits. Responsive models are simple, can be scaled to many agents,
and are easy to calibrate against data. But their agents have no objective, no veto, and no
strategy; therefore, they capture convergence rather than bargaining. Anticipatory models
represent vetoes and strategic choice directly, and they explain \emph{why} a proposal passes or
fails. But their equilibria become hard to compute as parties and issues multiply, and they are
rarely confronted with data from real negotiations.

Between these two groups lies a middle ground: models that retain the strategic, veto-bearing structure of the
second group but, instead of solving for equilibrium in closed form, learn it from repeated interaction. This
middle ground has grown into a field of its own and is treated separately in the next section.


\section{Negotiation and coalition formation simulation using reinforcement learning}
\label{sec:marl}

Section~\ref{sec:consensus} left two requirements for a model of ATS behaviour. The model must
represent many heterogeneous parties, each holding a veto and bargaining over a status quo. It
must also remain computable at that scale, which the closed-form equilibria of analytic game
theory are not. Multi-agent reinforcement learning (MARL) meets both requirements: it keeps the
game-theoretic substance but learns equilibrium behaviour from repeated interaction instead of
solving for it.

Formally, MARL is defined over stochastic (Markov) games, a generalisation of repeated games in
which the state of the negotiation changes as the agents act. Each agent observes the state,
chooses an action, and receives a reward: the objective function it learns to maximise. The
learning procedures approximate equilibrium behaviour over many simulated rounds rather than
deriving it in closed form \parencite{Littman1994}. Neural networks (NN) enter only as function
approximators, allowing this learning to scale to large state and action spaces.

Foundational MARL work covers multi-issue bargaining \parencite{Chang2021}, negotiating-team formation \parencite{Bachrach2020},
bargaining under risk and asymmetric information \parencite{Schmid2020}, and search-improved game-theoretic learning
\parencite{Li2023}. Two reviews cover the field: \textcite{Gronauer2022} on MARL and \textcite{Sarkar2022} on coalition formation. Most recently, MARL agents based on Large Language Models (LLMs) have been used to 
model political negotiation directly: for example, \textcite{Moghimifar2024} simulate coalition formation in parliament using text data
from party manifestos. \textcite{Bolleddu2025} present a MARL system for consensus building and negotiation based on synthetic data.
\textcite{Kulkarni2025} use negotiated agreements with game-theoretic reasoning to predict dynamic coalition formation in Diplomacy,
a board game in which seven powers negotiate, form, and break alliances, widely used as a benchmark for this type of simulation. Finally, some other benchmarks for real negotiation data are starting to appear 
\parencite{Benac2026}. However, these simulations run on synthetic data or abstract benchmarks rather than the deliberative records of a real consensus regime.


\section{Quantitative and text-based study of institutions and treaties}
\label{sec:computational}

The quantitative study of international institutions long predates its application to
Antarctica. Substantial literature codes the design of treaties and organisations as structured
variables: categorical or ordinal attributes such as their decision rules, dispute-settlement provisions, and enforcement mechanisms.
Some studies link these variables to institutional performance, survival, and decline
\parencite{Gray2018}. A more recent trend treats the \emph{text} of international agreements and debates as data \parencite{Grimmer2013}.
They recover state positions from voting records and speeches \parencite{Baturo2017,Slapin2008}. Together, these approaches
show that the preferences and design features of treaty regimes can be measured usefully, not merely
described: such measures track institutional outcomes and recover known positions.

The ATS's health has been assessed qualitatively \parencite{Mancilla2022} and quantitatively. Quantitatively, it
has been analysed in two ways. \textcite{Gardiner2025} provides performance accounts that measure the
regime in aggregate, counting inputs and outputs, and the time legally binding measures take to enter into force. It
concludes that responsiveness to challenges in the Antarctic region is declining.
A Working Paper (WP), the document type through which parties bring proposals to the ATCM
(Section~\ref{sec:ats}), may be submitted by a single party or co-sponsored by several. The second
methodology is structural: it uses this co-sponsorship to map who collaborates with whom,
through co-authorship networks of ATCM documents \parencite{Botelho_CoalitionDynamics}, and how such networks form \parencite{Kumar_ConsistentPatterns}.
This thesis uses both lines of work. The stances it reads from the text should explain the
co-sponsorship behaviour these networks describe: parties with similar stances are likely to
co-sponsor papers together, and Section~\ref{sec:ch2} turns this into a validation check against
observed co-sponsorship. The same measures should also help predict the regime's output series
reported by \textcite{Gardiner2025}, and Chapter 2 tests this.
Neither analysis, however, considers what the papers are \emph{about}: both count documents or
map their sponsors without reading the content. As a result, they cannot say which issues divide
the parties, a limitation taken up in Section~\ref{sec:gap}.


\section{The Antarctic Treaty System as a consensus regime}
\label{sec:ats}
At the ATCM, Consultative Parties (CPs) reach decisions exclusively by consensus; non-consultative parties (NCPs), observers, and experts attend but take no part. These outputs take three forms \parencite[Rule~24]{ATS2023RoP}. \textbf{Measures} set out
substantive provisions that the parties intend to be legally binding in their domestic law; they
require each party's government to formally approve them. \textbf{Decisions} settle the ATCM's own
internal and organisational matters. \textbf{Resolutions} are non-binding: they record
recommendations or expressions of intent, with no legal force.

There are four types of input:
\begin{itemize}
\item \textbf{Working Papers (WP)} \parencite[Rule~48]{ATS2023RoP}: submitted by CPs and Observers, these require
 discussion at the ATCM and are the main source of outputs. They are the only input from CPs that can produce an outcome.
\item \textbf{Secretariat Papers} \parencite[Rule~49]{ATS2023RoP}: organisational matters such as the budget.
\item \textbf{Information Papers} \parencite[Rule~50]{ATS2023RoP}: supporting information, also from NCPs and experts.
\item \textbf{Background Papers} \parencite[Rule~51]{ATS2023RoP}: documents submitted for the record only.
\end{itemize}

The consensus system makes the ATS a notable example of international cooperation \parencite{jackson2021antarctica}; however, the same mechanism
has become a major challenge as the number of CPs grows \parencite{Tsebelis2002,Gardiner2025}.

A notable case of contested consensus occurred in the late 1970s and 1980s, during the negotiations over mineral
resources in Antarctica. The Convention on the Regulation of Antarctic Mineral Resource Activities (CRAMRA) was
a framework, agreed in 1988, that would have permitted and regulated mining in Antarctica. It was
abandoned before entering into force and replaced by the Madrid Protocol (1991), which instead
prohibited all mineral resource extraction and designated Antarctica a natural reserve devoted to
peace and science \parencite{jackson2021antarctica}. Although it unfolded in two stages, it forms a single documented episode: one negotiation over
Antarctic mineral resources in which the parties' positions are recorded and then reversed. That
documented reversal makes it a valuable canary test, a known case against which any quantitative
stance measure can be validated.


\section{Measuring policy preference from political text}
\label{sec:computational-ir}

Positional measures place each actor on a policy scale, for example from favouring regulation
to favouring access, using only the text the actor produces. Extracting them from political
texts is a well-developed field with two paradigms. They differ in how they turn text into
numbers: the first paradigm uses word counts (i.e. frequencies), while the second uses embeddings.
The first paradigm is scaling and ideal-point methods, which recover a latent positional dimension
from the data itself: the axis is not chosen in advance, it emerges from patterns in the text, and
its meaning (e.g. regulation versus access) is interpreted after estimation. Wordfish \parencite{Slapin2008} is a Poisson scaling model of party text. Its aim was to
estimate party positions over time without hand-coded reference texts. The model treats each word
count as a Poisson draw whose expected frequency depends on the party's position on a single latent
dimension, and it recovered credible left--right trajectories for German parties from their
election manifestos (1990--2005). Its limitation is that the recovered dimension is uninterpreted:
the model finds the dominant axis of variation in word usage, which need not correspond to any
policy dimension of interest. In a corpus as topically diverse as the ATCM Working Papers, that
axis would likely reflect what parties talk about rather than what they support.
\textcite{Lauderdale2016} addressed the problem that legislators talk about different things in
different debates. Their model scales each debate separately with Wordfish and then combines the
debate-level positions into a general dimension through a Bayesian factor analysis. Applied to
speeches in the Irish D\'ail and the US Senate, it recovered within-party disagreement that voting
records hide. The approach still inherits the uninterpreted-dimension problem: each debate scale is
whatever axis dominates that debate's word frequencies, which need not be the substantive
for-or-against axis of interest. Lauderdale and Herzog get away with this because legislative
debates are adversarial, so the dominant axis is usually government versus opposition; ATCM Working
Papers have no such structure. Two things are therefore missing for this thesis: the topics have to
be discovered from the papers themselves rather than given in advance, and the output must be a
stance per topic rather than one general dimension.
\textcite{Vafa2020} produced a probabilistic framework that jointly places ideal points and models actors' language. The closest dataset to the ATCM corpus is the one \textcite{Baturo2017} use to derive
state-level preferences from United Nations (UN) debate speeches. Their corpus collects over 7300
annual General Debate statements (1970--2014), one per country per year, in a setting that
parallels the ATCM: formal, annual, state-level, and translated into official languages. They
derive preferences in two ways, both from word frequencies: Wordscores, a supervised scaling in
which the analyst defines the axis in advance through reference texts (US and Russian statements
anchor a US--Russia dimension), and correspondence analysis, an unsupervised reduction whose
dimensions are interpreted afterwards. Finally, \textcite{Parschan2025} review the text-based ideal-point estimation algorithms of this
first paradigm, comparing their assumptions and variants in one place.
The second paradigm uses embeddings as its main tool. An embedding maps a piece of text (a word,
a sentence, or a whole document) to a vector in $\mathbb{R}^d$, with $d$ typically in the hundreds.
The map is learned from large corpora so that texts appearing in similar contexts receive nearby
vectors; comparisons of meaning then become geometric operations, such as distances and angles
between vectors. Modern embeddings are produced by neural language models such as BERT
\parencite{Devlin2019}, which make the vectors context-dependent: the same word receives different
vectors in different sentences. \textcite{Rheault2019} show that word-embedding geometry recovers
ideological placement using parliamentary data. \textcite{Rodriguez2021} provide a comprehensive analysis of
what works and what does not when analysing political text using embedding techniques. They find embeddings useful for analysing political text, but caution that the choice of
embeddings strongly affects applied inference, and that the results must be validated.

This thesis adopts the second paradigm and uses embeddings at several stages: to discover what
topics the Working Papers discuss, to measure how relevant each paper is to each topic, and,
through the same family of neural models, to read each paper's stance on a topic. It also differs from the ideal-point methods in an important way. Ideal-point models summarise
each actor with one number: a single position on one latent axis, assumed stable across issues.
That summary is too coarse for the ATCM, where a party can, for example, support stricter
environmental protection yet oppose new tourism rules. This project therefore requires a
topic-conditioned stance: for each party, a separate position on each topic, allowed to change
through time.
The topic structure is discovered using BERTopic \parencite{Grootendorst_BERTopic}, which clusters
document embeddings and labels each cluster with its most distinctive words. The stance is read
with natural language inference (NLI) models \parencite{Burnham2023,Abercrombie2022}: classifiers
that take a text and a stated hypothesis (e.g. ``this paper supports stricter protection'') and
judge whether the text entails, contradicts, or is neutral towards it. The results can be stress-tested in three ways. The discovered topics are compared against the
categories that the Antarctic Treaty Secretariat assigns to each WP by hand. The stances are
validated against the CRAMRA canary test: applied to the papers of the minerals negotiations, the
pipeline should recover the documented positions, including the late reversal by Australia and
France \parencite{jackson2021antarctica}. Both can also be tested against observed co-sponsorship. This thesis assumes that countries
co-sponsor a WP because they share its topic and stance. That assumption is testable: within the
same year and topic, countries whose recovered stances are similar should be the ones that
actually co-sponsor papers, with stance similarity computed excluding the paper being predicted.
Section~\ref{sec:ch2} sets out these validation checks.


\section{Summary of literature gap}
\label{sec:gap}

The descriptive work, the aggregate performance accounts and qualitative health assessments of
Section~\ref{sec:computational} \parencite{Gardiner2025,Mancilla2022}, shows what the regime
outputs but not who produces what. The structural analyses of co-authorship networks
\parencite{Botelho_CoalitionDynamics,Kumar_ConsistentPatterns} are undirected and unattributed.
The text-based methods of Section~\ref{sec:computational-ir} recover latent positions rather than
topic-specific stances over time. Finally, consensus has been simulated either as convergence
(Section~\ref{sec:consensus}) or, where learning-based methods were used, on synthetic data rather
than the institution's own deliberative records (Section~\ref{sec:marl}). These literatures differ
in method but none recovers a directional preference: who is for or against what, extracted from
the regime's own words.

This gap has two parts. The knowledge gap is the question raised in Section~\ref{sec:lr-intro}:
whether the regime's outputs, including failures such as the 2022 ATCM's inability to agree its
final report, can be anticipated from the parties' positions at all. The methodological gap is
the missing instrument: no existing method measures each party's directional stance, per topic
and over time, from the regime's own records. Without that instrument the question cannot be
answered. Section~\ref{sec:rqs} turns this gap into three research questions.

No prior work extracts a directional preference for each party over time from a consensus
regime's own textual records, nor uses such preferences to drive a strategic simulation in which agents have veto
power. This thesis supplies both halves and joins them: the preferences recovered from the record become the
 parameters of the simulation that reproduces consensus and its failures.


\section{Research problem}
\label{sec:rqs}
The first goal of this thesis is to reconstruct parties' preferences from the written records 
into a directional measure (stance) across topics and time. The second goal is to extract
a directional opposition (veto) signal from the same records, and to test whether the combined 
preferences recover known history and predict decisions. Finally, this thesis aim to use these preferences
to parameterise a multi-agent model that simulates consensus and its failures.


