%% Section 04 — Program Design / Research Methodology
\chapter{Program and Design of the Research Investigation}
\label{ch:program}

\section{Research Questions}
\label{sec:rqs}
The thesis addresses the gap mentioned above by answering these reserch questions:
\begin{rqbox}
  \begin{enumerate}[label=\textbf{RQ\arabic*.}, leftmargin=*, itemsep=6pt]
    \item Can Parties’ strategic positions be reconstructed from written records into a directional
           measure across topics and time?
    \item Does the reconstructed positions recover known history and predict outcomes?
    \item Can a multi-agent model parameterised by it reproduce consensus and test counterfactual decision rules?
  \end{enumerate}
\end{rqbox}

\section{Objectives, Methodology and Research Plan}
\label{sec:objectives}

\subsection{Objective 1 / Chapter 1: The Support Layer}
\label{sec:ch1}

The \textbf{Aim} is to produce a defensible, directional measure of Parties' preferences over relevant topics through time, 
using documents the Parties submit themselves. For the ATCM case, Working Papers (WPs) are the input data because, of
all submitted documents, these are the only ones submitted by Parties or Observers (non-Secretariat) that require discussion and 
action. This makes WPs the most likely text to contain Parties' strategic intentions. The analysis uses English-language papers, since these
represent the largest share of all WPs. It is also relevant that most of the available pre-trained embedding and NLI models were trained
on English-language data. 

The \textbf{approach} is a pipeline with multiple stages. The first set of stages applies BERTopic \parencite{Grootendorst_BERTopic}
 to the whole corpus, producing a set of topics $K$ with corresponding cluster centroids $\{\boldsymbol{\mu}_k\}_{k \in K}$, where 
 each $\boldsymbol{\mu}_k \in \mathbb{R}^{m}$ in the $m$-dimensional embedding space. The second set of stages selects, for each 
 document and topic, the passages relevant to that topic. Each document $d$ is split into passages $P_d = \{p_1, \dots, p_{n_d}\}$,
 and each passage $p$ is embedded as $\mathbf{e}_p \in \mathbb{R}^{m}$. Relevance acts as a gating mechanism that filters out 
 irrelevant passages: for each document $d$ and topic $k$, every passage $p \in P_d$ is scored by its cosine similarity to the topic
centroid as follows,
\[
  \rho_{p,k} = \cos(\mathbf{e}_p, \boldsymbol{\mu}_k)
             = \frac{\mathbf{e}_p \cdot \boldsymbol{\mu}_k}
                    {\lVert \mathbf{e}_p \rVert \, \lVert \boldsymbol{\mu}_k \rVert}.
\]

The set of pairs $(document,topic)$ is the top scoring pairs paragraphs above a threshold. 

This may result in several $(document,topic)$ pairs for one document. Then looping over all these pairs, the stance on a 
named axis, is measured. These are then aggregated into a $Country \times Topics \times Year$ tensorr. 
 

\begin{equation}
S_{c,k,y} = \sum_{d \in D(c,y)} w_{d,c}\, r_{d,k}\, s_{d,k}
\end{equation}

Where $D(c,y)$ is the set of Working papers co-sponsored by country c in year y, $r_{d,k} \in [0,1]$ is the relevance of document $d$ to topic $k$, 
$s_{d,k} \in [-1,1]$ is the signed stance of 


\section{Research Design Overview}
\label{sec:design}

The project is structured into three interconnected papers, each targeting one research
question and one modelling contribution.

\begin{table}[htbp]
  \centering
  \caption{Paper structure and milestones}
  \label{tab:papers}
  \begin{tabular}{clll}
    \toprule
    \textbf{Paper} & \textbf{Focus} & \textbf{Method} & \textbf{Target venue} \\
    \midrule
    1 & Affinity dynamics     & Network analysis, coalitional GT    & TBD \\
    2 & Aversion dynamics     & Veto-player model, ABM              & TBD \\
    3 & Simulation framework  & Agent-based model, RL               & TBD \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Paper 1 — Affinity Dynamics in Multilateral Treaties}
\label{sec:paper1}

% TODO: ~500 words.
%       - Formal definition of affinity between treaty parties
%       - Data sources (ATS meeting records, voting data)
%       - Network construction and community detection
%       - Expected contribution: first systematic map of affinity coalitions in ATS

\section{Paper 2 — Aversion Dynamics and Consensus Breakdown}
\label{sec:paper2}

% TODO: ~500 words.
%       - Formal definition of aversion / veto propensity
%       - Game-theoretic model of strategic blocking
%       - Calibration to historical ATS consensus failures
%       - Expected contribution: predictive model of consensus breakdown conditions

\section{Paper 3 — Integrated Simulation Framework}
\label{sec:paper3}

% TODO: ~500 words.
%       - Architecture of the agent-based model (state agents, treaty rules, environment)
%       - RL-trained agents for adaptive strategy
%       - Validation against historical ATS outcomes
%       - Generalisability tests on other treaty systems (climate, fisheries)

\section{Data Sources and Ethics}
\label{sec:data}

The primary data sources for this project are:

\begin{itemize}
  \item Antarctic Treaty Consultative Meeting (ATCM) final reports and working papers
        (publicly available at \url{https://www.ats.aq/}).
  \item State-level foreign policy position datasets (e.g., UN voting records).
  \item Published network and voting data from the international relations literature.
\end{itemize}

All data are publicly available institutional records; no human-subjects research is
involved and no ethics clearance is required beyond standard QUT research integrity
protocols.

\section{Timeline}
\label{sec:timeline}

\begin{longtable}{lll}
  \caption{Indicative project timeline} \label{tab:timeline} \\
  \toprule
  \textbf{Phase} & \textbf{Activity} & \textbf{Target Date} \\
  \midrule
  \endfirsthead
  \toprule
  \textbf{Phase} & \textbf{Activity} & \textbf{Target Date} \\
  \midrule
  \endhead
  \midrule
  \multicolumn{3}{r}{\small Continued on next page} \\
  \endfoot
  \bottomrule
  \endlastfoot
  1  & Literature review \& theoretical framework      & Month 6  \\
  2  & Paper 1: data collection \& analysis            & Month 12 \\
  3  & Paper 1: writing \& submission                  & Month 15 \\
  4  & Paper 2: model development \& calibration       & Month 18 \\
  5  & Paper 2: writing \& submission                  & Month 21 \\
  6  & Paper 3: simulation framework development       & Month 27 \\
  7  & Paper 3: writing \& submission                  & Month 30 \\
  8  & Thesis integration \& write-up                  & Month 36 \\
  9  & Thesis submission                               & Month 38 \\
\end{longtable}

\section{Expected Contributions}
\label{sec:contributions}

\begin{enumerate}
  \item A formal computational model of consensus dynamics in the ATS, the first of
        its kind at this level of institutional specificity.
  \item A transferable agent-based simulation framework applicable to any
        consensus-based multilateral treaty.
  \item New theoretical constructs (affinity/aversion indices) operationalising
        previously qualitative political science concepts.
  \item Policy-relevant insights into treaty design features that improve stability
        under strategic pressure.
\end{enumerate}
