%% Section 06 — Appendices
\chapter{Appendices}

\section*{Appendix A: Research Integrity Online}
\addcontentsline{toc}{section}{Appendix A: Research Integrity Online}

[Insert RIO completion certificate. I am on it, next I will add it to this document before submission]

\section*{Appendix B: Coursework}
\addcontentsline{toc}{section}{Appendix B: Coursework}

AIRS / IFN006 is underway, this course will end on july.

\section*{Appendix C: Notation}
\addcontentsline{toc}{section}{Appendix C: Notation}

\paragraph{Indices and sets.}
\begin{tabular}{@{}ll@{}}
$c, i$        & countries (Consultative Parties), $c \in \{1,\dots,C\}$ \\
$k$           & topic, $k \in K$; $|K|$ is the number of topics \\
$y$           & year \\
$d$           & document (Working Paper) \\
$p$           & passage (chunk) of a document \\
$\ell$        & ordinal stance class, $\ell \in \{1,\dots,5\}$ \\
$D(c,y)$      & Working Papers co-sponsored by country $c$ in year $y$ \\
$K(d)$        & topics assigned to document $d$ \\
$P_d$         & passages of document $d$, $P_d=\{p_1,\dots,p_{n_d}\}$ \\
$P_{d,k}$     & passages of $d$ selected as relevant to topic $k$ \\
$A_d$         & co-sponsors of document $d$ \\
\end{tabular}

\paragraph{Embedding space.}
\begin{tabular}{@{}ll@{}}
$m$                  & embedding dimensionality (full Stella space) \\
$\mathbf{e}_p$       & embedding of passage $p$, $\mathbf{e}_p \in \mathbb{R}^{m}$ \\
$\boldsymbol{\mu}_k$ & centroid of topic $k$, $\boldsymbol{\mu}_k \in \mathbb{R}^{m}$ \\
$\rho_{p,k}$         & passage--topic relevance, $\rho_{p,k}=\cos(\mathbf{e}_p,\boldsymbol{\mu}_k)$ \\
$\tau$               & relevance selection threshold \\
\end{tabular}

\paragraph{Stance and hypotheses.}
\begin{tabular}{@{}ll@{}}
$h_k^{+}, h_k^{0}, h_k^{-}$ & for-, neutral-, against-pole hypotheses for topic $k$ \\
$\lambda(d,h)$  & null-premise-debiased entailment log-ratio (DCPMI) \\
$\tilde\pi_{d,k}$ & three-pole distribution (stance classifier) \\
$\pi_{d,k}$     & five-class ordinal distribution, $\pi_{d,k}\in\Delta_5$ \\
$\Delta_5$      & probability simplex, $\{\pi\in\mathbb{R}^5 : \pi_\ell\ge 0,\ \sum_\ell \pi_\ell = 1\}$ \\
$v$             & class values, $v=(-1,-\tfrac12,0,\tfrac12,1)$ \\
$s_{d,k}$       & signed stance of $d$ on topic $k$, $s_{d,k}\in[-1,1]$ \\
\end{tabular}

\paragraph{Aggregation and tensors.}
\begin{tabular}{@{}ll@{}}
$w_{d,c}$           & attribution weight: $1$ (full credit) or $1/|A_d|$ (fractional) \\
$S_{c,k,y}$         & support of country $c$ on topic $k$ in year $y$, Eq.~(\ref{eq:aggregate}) \\
$S_{\cdot,\cdot,y}$ & year-$y$ stance slice, $S_{\cdot,\cdot,y}\in\mathbb{R}^{C\times|K|}$ \\
$O_{c,k,y}$         & opposition of country $c$ on topic $k$ in year $y$ \\
\end{tabular}

For these tensors, high sparsity is expected because absence of evidence is recorded as missing, not as zero.

\paragraph{Stance classifier (Chapter~1).}
For topic $k$, let $h_k^{+}, h_k^{0}, h_k^{-}$ be the for-, neutral-, and
against-pole hypotheses. For passage set $d$ and hypothesis $h$, the debiased evidence is the null-premise-corrected log-ratio,
\[
  \lambda(d,h) = \operatorname{clip}\!\Big(
     \log P(\text{entail}\mid d,h) - \log P(\text{entail}\mid \varnothing,h),
     \ \pm 5 \Big),
\]
this removes each hypothesis's marginal acceptability. The three-pole distribution
combines entailment of each pole with contradiction of its opposite,
\[
  \tilde\pi_{d,k} = \operatorname{softmax}\!\begin{pmatrix}
     \lambda_{\mathrm{ent}}(h_k^{+}) + \lambda_{\mathrm{con}}(h_k^{-}) \\[2pt]
     \lambda_{\mathrm{ent}}(h_k^{0}) \\[2pt]
     \lambda_{\mathrm{ent}}(h_k^{-}) + \lambda_{\mathrm{con}}(h_k^{+})
  \end{pmatrix},
\]
these values are expanded to the five ordinal classes $\pi_{d,k}$ and reduced to the
scalar stance by the expected value of Equation~\ref{eq:stance}.

\section*{Appendix D: Acknowledgement of AI Use}
\addcontentsline{toc}{section}{Appendix D: Acknowledgement of AI Use}
I have used Claude Code for boilerplate and scaffolding code in the implementation of the pipelines and scripts.
All architectural decisions, methodological choices, and algorithm design were specified by me in the input prompts. I also
reviewed and edited generated code. 
I have also used Consensus AI and the Claude app to fetch the most relevant scientific papers from the available literature, based on my own
ideas, keywords, and criteria. Afterwards, I read the papers and drew conclusions myself. 
Finally, I used Claude Code as a writing assistant for this document, to help with grammar, readability, clarity and LaTeX document formatting. All research questions, 
project design, ideas, interpretation of results, and selection of literature are my own.
